A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data (Hadoop Or Not Hadoop)

### 2014-03-12 paper-plos.tex
Extensive changes all over

\usepackage{color}

@@

\definecolor{ChangeColor}{rgb}{0.24,0.56,0.18}
\newcommand{\Change}[1]{{\color{ChangeColor}{#1}}}

\definecolor{FixmeColor}{rgb}{1,0,0}
\newcommand{\Fixme}[1]{{\color{FixmeColor}{#1}}}

@@

%Current practices: How to parallelize
The most common approach to accelerate NGS tools is to parallelize within a compute node (multi-core parallelization) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16 (see~\cite{top500}). For the tools which do not support OMP natively, e.g. Samtools, variant calling can be parallelized by creating a separate process for each chromosome or using the GNU Parallel Linux utility~\cite{Tange2011a}.
It is important to note that a multi-core approach does not improve the performance of operations that are limited by local disk or network throughputs, whereas splitting the dataset and using multi-node parallelization is not generally bound by those constraints.

@@

@@

We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data, one based on Hadoop and the other on regular HPC with a batch processing system (hereafter referred to as the HPC method) (Methods).
Our experiments consisted of executing the pipelines for five datasets with short reads of different sizes and measuring the wall-clock run time for each pipeline stage.
All experiments were repeated several times and the results averaged to obtain a data point for the particular combination of data size and computational platform.
Finally, we propose modifications for existing Hadoop software to better benefit fully from massively parallel nature of MR computations; we also
developed a set of Bash scripts for the classical HPC DNA-seq analysis programs for short read alignment to use multiple nodes and fully exploit the network, thereby speeding up calculations.

\subsection*{Modified preprocessing stage}
A native preprocessing stage in Crossbow provides great flexibility in delivering the data to the Hadoop cluster, where data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet~\cite{Langmead:2009kx}. However, the only way to introduce parallelization in the native preprocessing stage is to split the read files into smaller chunks and to explicitly configure Crossbow's preprocessor to download each chunk (this is done by listing them in a special \textit{manifest file}. listing all these chunks. This, however, is lacking the massively parallel way of treating the data.

%\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis}

Assuming that sequencing platforms are in close proximity of storage and
computing facilities, we considered that the data was already stored where
Hadoop could access it.
We rewrote the Crossbow preprocessing stage scripts as MR programs, thus
benefiting from its massively parallel nature.  For our tests, we required that
the input FASTQ data be compressed with \texttt{bzip2}.
\texttt{bzip2} provides very good compression and is splittable,
While this is not an
intrinsic constraint in our set up, we reasoned that it is typical to store
sequencing data in a compressed file format; \texttt{bzip2} provides very good
compression and its file format is splittable,
meaning that the archive can be expanded in a parallel manner allowing Hadoop to process multiple segments simultaneously.
In addition, the \texttt{bzip2} format is natively supported by Hadoop -- i.e. FASTQ data can be read whether it is uncompressed or \texttt{bzip2}-compressed.

In our tests, the storage volume containing the input data was mounted with SSHFS to one of the Hadoop nodes, from where the \texttt{bzip2}-compressed data were transferred to the HDFS.

@@

The Hadoop streaming library offers the flexibility to write Hadoop programs in any programming language. Our Python scripts efficiently process short reads, produced by Illumina sequencing platform of different versions, as well as the FASTQ files converted from SRA format (NCBI Sequence Read Archive~\cite{ncbi-sra}).
% LP: I think this sentence could be removed since to appreciate what is being
% described one would have to understand quite well how that program works.
The various formats that our script reads lack a standard FASTQ record id, which needs to be interpreted properly at least to extract the correct read number.  To circumvent this problem we rewrote the id based on a SHA-1 hash function and labelling reads mates ``.1" and ``.2" suffixes.
To ensure that both the forward and reverse reads of the same pair would end up on the same Reducer, secondary Mapper key-sorting mechanisms were involved.

In order to compare the native preprocessor and the proposed one, the data was put on HDFS beforehand to eliminate possible network delays -- e.g., while downloading data from Amazon.
Table~\ref{table:preprocess} illustrates the benefits of our MR approach.
To complete the comparison the same preprocessing was performed with a BASH script for the same data located on the HPC storage. The script utilizes multiple cores and was executed on a single HPC node (see Supplementary information).
%ran on a single HPC node, utilizes multiple cores, with the limiting factor being \sout{the hard disk IO performance.}\COM{More correctly to say, that the pipe is not filled completely. i.e. the degree of achieved parallelization does not utilize all 16 cores. Perhaps not the best bash commands etc.}
%, which effectively leaves with 5-7 cores of 16. \COMMENT{awkward}

\subsection*{Accuracy of pipelines}
Since our HPC and Hadoop pipelines use different~SNP callers (Samtools and SOAPsnp~\cite{soapsnp}, respectively) we do not expect them to deliver perfectly matching SNP lists, but we do expect them to capture and correctly identify the
% which mutation?
\Fixme{mutations}.
We only tested for correctness with the smallest dataset~(I). The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$~\cite{schneeberger} was successfully localized by both variant callers.

\subsection*{Scalability of HPC and Hadoop approaches}
\Change{To demonstrate the scalability of the HPC approach we ran the HPC pipeline using dataset~I as input. We collected the running times} as a function of the number of cores used (Table~\ref{table:3}).
We found that the alignment process with Bowtie scales fairly well, while SNP calling with Samtools is the bottleneck of the workflow. Indeed, this latter part scales worse than Bowtie and consumes a progressively larger portion of the total calculation time as the input grows, even after parallelizing analysis by chromosome as previously suggested~\cite{biostars_samtools}.
%\footnote{\url{http://www.biostars.org/p/48781}
%\COM{here was a footnote referring to biostars site. Ok, if the footnotes are not welcome, lets then have it as a reference. That code is not trivial. Maybe we can leave the ref?}
%\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance. During its sorting phase the lack of RAM results in spilling to the HDD. For example for an half of  Dataset III, SNP calling takes approximately 40 minutes with 8 GB RAM, and 15 minutes with 48 GB RAM, when  the whole BAM file fits into RAM. For our tasks we used the default Samtool's settings.}.
\Change{For both the HPC and Hadoop approaches, and using all the test datasets (I-V), we collected the time required for the combined task of mapping the reads to the corresponding reference genome and SNP calling. The results are shown in Table~\ref{table:4}.}

\Change{The results confirm that one the advantages to using the Hadoop~MR} framework is its almost linear scalability -- i.e.\ the calculation time linearly depends on the size of the dataset~\cite{Langmead:2009kx,Pireddu:2011vn}. \Fixme{Deleting this since one could dream/find pathological cases that don't work: ``regardless the scaling nature of the underlying program''.}
Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for the 56-core Hadoop cluster and datasets I-IV.
%To stress the linear nature of scaling, both sets of points were fit using the least squares method.

\subsection*{Comparing Hadoop and HPC runtime efficiency}
\Change{%
Hadoop was designed to process huge datasets\cite{hadoop,lin2010}. Therefore, one could image then that the larger the dataset the more advantageous Hadoop should become compared to the HPC approach. In order to compare the efficiency of the two processing platforms (Hadoop and HPC) as the size of the dataset varied, we looked at the number of core-hours used for same analysis:
$F=T_{p}\times p,$ 
where~$T_{p}$ is the calculation time on~$p$ cores. 
}

\Change{%
Using the data from Table~\ref{table:4}, in Figure~\ref{fig:fig2} we plot the ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size. Extrapolation to zero on the~$X$-axis estimates the ratio for the hypothetical infinite dataset. It is evident from the data that the Hadoop pipeline becomes increasingly more efficient than the HPC-based pipeline as the dataset size increases. 
The parabolic extrapolation of our data yields the value~$1.70\pm0.01$ for the ratio, meaning that Hadoop, even if running in the virtualized environment of a private cloud running on modest hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
}

\Change{%
The simplest curve that fits the data points nicely is a parabolic function which, importantly, is not a constant. 
Since our measurements using the Hadoop-based pipeline, summmarized in Figure~\ref{fig:fig1}, showed linear scaling properties of Hadoop in this setting, a deviation of the ratio $F_{Hadoop}/F_{HPC}$ from the constant is evidence of the 
lack of scalability of the HPC approach.

%Consider the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
%\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e. being a constant; thus the ratio of two constants would result a constant, but we observe a dependecy.}. \COMMENT{Why jump back and forth in the discussion of figures 1 and 2?} 
We suspect the main reason for such a behavior is that for large datasets (dataset~III-V, in our case) Samtools' SNP calling routine becomes more memory- and time-demanding than the alignment with Bowtie.
Since its memory requirements scale with input size, with sufficient input data Samtools swaps data to disk, creating multiple (up to hundreds) temporary files while it sorts the BAM file -- incidentally, these files can end up on shared network-accessed file systems and thus generate large amounts of network traffic.
Table~\ref{table:5} shows the time required by Samtools to sort dataset~IV with different amount of RAM available on a single HPC node. Though there is a large deviation when using small amounts of RAM, we believe due to other users' load on the cluster network, the trend quite clearly indicates that a relatively large amount of RAM per node is needed in order to use Samtools effectively. In contrast, in our tests Hadoop performed the SNP calling on much more economic hardware in linear time with just 12GB of RAM per \textit{virtual} node.
}

\subsection*{Comparing the network communication efficiency for the Hadoop and HPC approaches }
The network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages. The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance depletion.
To study this phenomena, we compared HPC and Hadoop network communication costs depending on the number of nodes involved for a fixed dataset size (58~Gbases, dataset~IV).

Due to the trivially parallelizability of the alignment process~-- the read-pairs are independent of each other, and can be aligned independently~-- one could try to involve more computational resources, e.g., split the initial data into chunks to process them independently.
Reducing the size of each data chunk reduces the aligner  job,~$T_{alignment}$, but at the same time, the more chunks almost simultaneously have to be sent over the network, potentially causing traffic jams, and therefore increasing the communication costs,~$T_{comm}$.

There are several program packages for short read alignment with MPI support~\cite{pmap, gnumap}, and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}. However, e.g., the Pmap package functions poorly for datasets larger than 20 Gbases. To circumvent this limitation we implemented a highly optimized Bash script making use of standard Unix utilities to use the HPC cluster network as efficiently as possible~\cite{code_repo_bash}, and compared the network performance with the standard Hadoop HDFS approach. We separated the alignment time, $T_{alignment}$, and the communication time, $T_{comm}$, and plotted $T_{alignment}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$ (Figure~\ref{fig:fig3}). 
This measure is applicable to both HPC and Hadoop, however, $T_{comm}$ has a different explanation. For Hadoop, the short reads in FASTQ format have to be preprocessed (involving node communication~$T_{comm}$) to be able to run in MR-fashion, while the data locality will be automatically achieved during the data ingestion into the HDFS. 
We rewrote the code for the preprocessing stage for Crossbow to make it suitable for MR-style parallelization.
For the HPC approach, $T_{comm}$ involves the chunks that are sent from the sequence delivery location to the local node scratch disks where the actual alignment happens, and the sending of the aligned SAM files back to the delivery location over the network.
%\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}}

@@

%The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}   is as follows: 
\textit{HPC results (open circles in Figure~\ref{fig:fig3})}: The data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratch disks of the nodes allocated by SLURM. One can see two distinct linear stretches. One stretch is for the range from 4 to about 12 nodes, and the another is from 12 up to 60. The former (horizontal stretch) is explained as for Hadoop~-- the more nodes involved, the faster the chunks are being distributed. The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated
%\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbasesit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.} 
and unable to pass more data in a time unit, while the alignment time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{alignment}\propto D\propto 1/N \rightarrow T_{alignment}/T_{comm}\propto 1/N$, i.e., a linear dependency, which can be observed in the plot. 
The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec. Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch. The nodes are being allocated on the same rack, which is the default SLURM behavior.

@@

In this report we have described two approaches for high-performance analysis of DNA sequencing data; one based on regular HPC using a batch system and one based on Hadoop. We show that Hadoop installed on a private cloud is an appealing solution both on terms of runtime performance and also in terms of usability. A key result is the data size where Hadoop is favorable to regular HPC batch systems, and in order to compare them we developed highly optimized pipelines for both scenarios. We found that dataset sizes larger than 100\,Gbases is where Hadoop excels over HPC, and also that Hadoop shows almost linear scalability for this problem. Extrapolation to infinite dataset size reveals however that HPC provides the results faster, given the same amount of resources as Hadoop on a private cloud.

>>>>

% required by the coloured \Fixme and \Change macros
\usepackage{color}

@@

\definecolor{ChangeColor}{rgb}{0.24,0.56,0.18}
%\definecolor{ChangeColor}{rgb}{0,0,0}
\newcommand{\Change}[1]{{\color{ChangeColor}{#1}}}

\definecolor{FixmeColor}{rgb}{1,0,0}
%\definecolor{FixmeColor}{rgb}{0,0,0}
\newcommand{\Fixme}[1]{{\color{FixmeColor}{#1}}}

@@

%Current practices: How to parallelize
The most common approach to accelerate NGS tools is to parallelize within a compute node (multi-core parallelization) using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16 (see~\cite{top500}). For the tools which do not support OMP natively -- e.g. Samtools -- variant calling can be parallelized by creating a separate process for each chromosome or using the GNU Parallel Linux utility~\cite{Tange2011a}.
It is important to note that a multi-core approach does not improve the performance of operations that are limited by local disk or network throughputs, whereas splitting the dataset and using multi-node parallelization is not generally bound by those constraints.

@@

\Fixme{move this section back to the its original position if you don't like it
here (it was just before Acknowledgements)}

\subsection*{Analysis pipelines}

For our work to determine when Hadoop is an appealing option for genomic
sequence analysis, we chose as a representative 
use case for our experiments the task identifying
single-nucleotide polymorphisms (SNPs) from short-read data.  We constructed
two pipelines that perform the same task: one that runs in an HPC environment
and the other on the Hadoop platform.  Although there are many different
approaches and software tools to perform this task in an HPC environment (e.g.,
GATK~\cite{gatk}, \Fixme{Other example? Otherwise none}), there is only a limited
number of options for Hadoop.  Given this limitation, we decided to keep the
analysis pipeline as simple as possible to ensure a fair comparison between the
two platforms.

\begin{enumerate}
\item HPC approach
\subitem Short-read alignment: Bowtie~ver.~0.12.8
\subitem SNP calling: Samtools~ver.~0.1.19
\item Hadoop approach: Crossbow
\subitem Short-read alignment: Bowtie~ver.~0.12.8
\subitem SNP calling: SOAPsnp~1.02
\end{enumerate}

In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the aligned reads and~SNP calling with Samtools. The Bowtie aligner natively implements~OMP -- therefore, using 8 cores on the same computer will theoretically produce a result 8 times faster than using a single core. Likewise, Samtools (as of version~0.1.19) also offers shared memory parallelism for several of its functions. Where available these features were used to improve the analysis speed. The exact workflow used is available from GitHub~\cite{code_repo_bash}.

\Change{%
The equivalent Hadoop-based pipeline was implemented with Crossbow. The input
data and the indexed genome reference were copied to Hadoop's storage
system~(HDFS) before starting the experiments.  Specifically, the storage volume
containing the input data was mounted with SSHFS to one of the Hadoop nodes,
from where the \texttt{bzip2}-compressed data were transferred to the HDFS\@.
This preparation was done to eliminate unpredictable effects on the experiments
from possible network delays.
}
Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.  Unfortunately, unlike Crossbow's main program, its preprocessor does not scale well to large datasets (it is not implemented as a MR program). Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments. To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation (available from GitHub~\cite{code_repo_mr}). Our implementation is described in Section~\ref{sec:preprocessor}.

\subsection*{Datasets}
For our experiments we used three publicly available DNA-seq datasets~(I--III),
a synthetic dataset~(IV) of {\it A.thaliana} -- the well-known model plant --
and a dataset (V) of two {\it H.sapiens} individuals
(Table~\ref{table:datasets}). Datasets I-III and V are composed of Illumina
HiSeq sequencing data. Additional information about the source of the datasets is
provided in the Supplementary Material section.

\Change{%
All data was standardized to FASTQ format and compressed with \texttt{bzip2},
since it is typical to store sequencing data in a compressed file format. The
\texttt{bzip2} compression codec was selected it provides very good compression
and is splittable by Hadoop, meaning that the archive can be expanded in a
parallel manner allowing Hadoop to process multiple segments simultaneously.
Also, since some of the datasets lack a standard FASTQ record id, which is
required by our test pipelines to extract the read number, we also standardised
the read ids in the datasets by generating new ids based on the SHA-1 hash
function and labelled read mates with ``.1" and ``.2" suffixes.
}

\subsection*{MapReduce-based Data Preprocessor}
\label{sec:preprocessor}

\Change{%
The Crossbow package includes a native preprocessing stage that provides great
flexibility in delivering the data to the Hadoop cluster; it can autonomously
download data from Amazon S3, FTP and HTTP servers over the
Internet~\cite{Langmead:2009kx}. However, this preprocessor does not parallelize
its operations. In theory one can manually introduce some degree of
parallelization by manually splitting the original input read data into smaller
files and to explicitly configuring Crossbow's preprocessor to download each one
(this is done by listing all of them in a special \textit{manifest file}).
Nevertheless, in practice this approach is rather error-prone and laborious.

%\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis}
We reimplemented the main functionality in the Crossbow preprocessor as an MR
program to benefit from its massively parallel nature. The new MR preprocessor
was implemented in Python using the Hadoop streaming library.
\Fixme{Consider deleting the following implementation detail.}
To route both reads of the same pair to the same Reducer, a secondary
key-sorting mechanism was employed.
The resulting program efficiently processes short reads produced by different
versions of the Illumina sequencing platform as well as FASTQ files generated
from SRA files (NCBI Sequence Read Archive~\cite{ncbi-sra}). However, the
Hadoop-based approach imposes the requirement that the input data be stored
where the Hadoop nodes can access it, which is in our opinion a reasonable since
sequencing platforms generally have direct access to extensive storage and
computing facilities.
}

\subsection*{Computational resources}
To run the HPC analysis pipeline we used a computational cluster at UPPMAX,
which is heavily used for analysis of NGS-data~\cite{lampa}.
Data and reference genomes were stored on a parallel, shared storage system.
On the other hand, the Hadoop test platform was deployed on a private cloud at
UPPMAX using the OpenNebula~\cite{opennebula} cloud manager. The cluster was
configured with Cloudera Hadoop distribution version~2.0.0-mr1-cdh4.5.0~\cite{cloudera}.
For the fully detailed hardware configuration of the two computational resources
please see the Supplementary Material section.

@@

\Change{%
As described in the Methods Section, we constructed two pipelines for identifying single-nucleotide polymorphisms
(SNPs) from short-read data, one based on Hadoop and the other on regular HPC
with a batch processing system (hereafter referred to as the HPC method)
(Methods).  Our experiments consisted of executing the pipelines for each of the
five input 
datasets (see Table~\ref{table:datasets}) and
measuring the wall-clock run time for each pipeline stage.  All experiments were
repeated
\Fixme{several times (how many?)}
and the results averaged to obtain a data point for the
particular combination of data size and computational platform.
}

\Fixme{Belongs with Discussion:
In light of our observations, we
propose modifications for existing Hadoop software to better exploit the
massively parallel nature of MR computations. Finally, we developed a set of Bash
scripts to parallelize the operations of classical DNA-seq analysis programs for
short-read alignment over multiple nodes and fully exploit the network bandwidth, thereby
accelerating calculations.
}

\subsection*{Accuracy of Pipelines}
\Change{%
Since our HPC and Hadoop pipelines use different~SNP callers (Samtools and
SOAPsnp~\cite{soapsnp}, respectively) we could not expect them to deliver perfectly
matching SNP lists. Instead, we verified the accuracy of the test pipelines by
verifying their ability to correctly identify the mutation $C\rightarrow T$ on
chromosome 4 at position $16702262$~\cite{schneeberger} in Dataset I.
}

\subsection*{Performance evaluation of MR-based preprocessing stage}

\Change{%
To quantify the processing time saved by adopting the MR approach for
preprocessing the read data, we ran an
experiment comparing the new preprocessor described in
Section~\ref{sec:preprocessor}  to the original preprocessor furnished with
Crossbow.  To further validate our strategy, we also compared the performance of
our MapReduce approach to a different parallelization strategy based on a custom Bash script
that uses of all processing cores available on a single computing node to accelerate the operation (see
Supplementary Material for details of the implementation). This program read its
input directly from conventional HPC storage, rather than HDFS\@.  The experiment
was run on Datasets II, III, and IV\@.  The Hadoop cluster used was configured
with 56 cores (7 slave nodes) while the Bash approach ran on a single node
equipped with 16 CPU cores.
}

@@

\Fixme{Alexey, please verify whether my interpretation seems reasonable. LP}

\Change{%
Table~\ref{table:preprocess} shows the results of the experiment. Clearly,
either of the parallel approaches are much faster than the original tool, with
all dataset sizes.  Also, we noticed that the difference between the Bash and Hadoop
approaches is minimal when running on Dataset II\@.  Indeed, with the relatively
small dataset size the overhead incurred by the Hadoop framework likely
overwhelms the useful compute time used for each available CPU core. However,
with increasing input size the superior scalability of MR becomes apparent, with
the largest dataset taking requiring less than half the time of the Bash
approach to be processed; at the same time, it is important to also notice that
the Hadoop approach employed approx.\ 1.6 times more CPU time compared to the
Bash approach.
}

%ran on a single HPC node, utilizes multiple cores, with the limiting factor
%being \sout{the hard disk IO performance.}\COM{More correctly to say, that the
%pipe is not filled completely. i.e. the degree of achieved parallelization does
%not utilize all 16 cores. Perhaps not the best bash commands etc.} , which
%effectively leaves with 5-7 cores of 16. \COMMENT{awkward}

\subsection*{Scalability of HPC and Hadoop approaches}
\Change{%
Given the wide range in genome dataset sizes processed in practice -- e.g.,
exomes,targeted sequencing, whole-genome sequencing at various depths -- it is
important to understand the scaling characteristics of each processing
platforms (Hadoop and HPC) with respect to input size. For each platform, we
measured the time required for to execute their respective variant calling
pipeline (see Methods) for each of the test datasets (I-V). The results are
shown in Table~\ref{table:4}.

The measurements highlight that one of the advantages of using the Hadoop~MR} framework is
its almost linear scalability with respect to input size -- i.e.\ the
calculation time linearly depends on the size of the
dataset~\cite{Langmead:2009kx,Pireddu:2011vn}.
%Deleting this since one could dream/find pathological cases that don't work: ``
\Fixme{\sout{regardless the scaling nature of the underlying program.}}
Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the wall clock
running time as a function of the dataset size for the 56-core Hadoop cluster and datasets
I-IV.
%To stress the linear nature of scaling, both sets of points were fit using the
%least squares method.

\Change{%
Analogously, we investigated the scalability characteristics of the HPC method
with respect to input size.
\Fixme{The following works better the aid of a plot.}
Analyzing the observations in Table~\ref{table:4}, we can see with the HPC
method run time increases superlinearly as a function of dataset size.  As
expected, the result is that the HPC method is not able to complete the analysis
with our larger tests datasets.  On the other hand, this approach is not much
slower than our Hadoop cluster considering the much smaller quantity of computing
resources that is uses -- an issue that is analyzed more closely in the section
on the efficiency of the two systems.
}

\Fixme{The following paragraph comments on how the HPC approaches' runtime
varies with the number of cores available, rather than input size. If we're
going to leave it in, I think we should make an analogous discussion regarding
how Hadoop's runtime varies with increasing resources. The data in Table 6 might
be suitable for this.}

To investigate the scalability characteristics of the HPC approach we ran the
HPC pipeline using Dataset~I as input. We collected the running times as a
function of the number of cores used (Table~\ref{table:3}). The results show that the
alignment process with Bowtie scales fairly well, but overall the pipeline
suffers because of the SNP calling step with Samtools. Indeed, this operation
is the bottleneck of the workflow as it scales
worse than Bowtie and consumes a progressively larger portion of the total
calculation time as the input grows -- even after parallelizing analysis by
chromosome as previously suggested~\cite{biostars_samtools}.

%\footnote{\url{http://www.biostars.org/p/48781} \COM{here was a footnote
%referring to biostars site. Ok, if the footnotes are not welcome, lets then
%have it as a reference. That code is not trivial. Maybe we can leave the ref?}
%\footnote{The amount of available memory, specified with the \texttt{-m}
%option, is of big importance for the Samtools performance. During its sorting
%phase the lack of RAM results in spilling to the HDD. For example for an half
%of  Dataset III, SNP calling takes approximately 40 minutes with 8 GB RAM, and
%15 minutes with 48 GB RAM, when  the whole BAM file fits into RAM. For our
%tasks we used the default Samtool's settings.}.

\subsection*{Comparing the efficiency of Hadoop and HPC}
\Change{%
Hadoop was designed to process huge datasets by distributing computations to a
large number of computers\cite{hadoop,lin2010}.  Therefore, while allowing us to
process very large datasets, the mechanisms required for distributing the work
add a non-trivial overhead to the computation, which becomes more and more
significant as the amount of work to be done is reduced -- i.e.\ the dataset
size decreases.

In order to measure the overhead incurred by the Hadoop platform, we compared
the efficiency of the two processing methods (Hadoop and HPC) as the size of the
dataset varies. We measure the efficiency indirectly as the number of core-hours
($F$) used for same analysis: $F=T_{p}\times p,$ where~$T_{p}$ is the running time on~$p$ cores.
}

\Change{%
\Fixme{The way of reorganized this text, at the end of the subsection it kind of
excessively teeter-totters too much between HPC and Hadoop.}

Using the data from Table~\ref{table:4}, in Figure~\ref{fig:fig2} we plot the
ratio $F_{Hadoop}/F_{HPC}$ as a function of the reciprocal dataset size.
%
The simplest curve that fits the data points nicely is a parabolic function
which, importantly, is not a constant.  Since our measurements using the
Hadoop-based pipeline, summarized in Figure~\ref{fig:fig1}, showed linear
scaling properties of Hadoop in this setting, a deviation of the ratio
$F_{Hadoop}/F_{HPC}$ from the constant is evidence of the lack of scalability of
the HPC approach.
}%
%Consider the linear scaling for the Hadoop approach for the same datasets, we conclude that HPC approach scales worse than linearly.
%\footnote{For the linear scaling the curve of $F=T_{p}\times p$ does not depend on the dataset size, i.e. being a constant; thus the ratio of two constants would result a constant, but we observe a dependecy.}. \COMMENT{Why jump back and forth in the discussion of figures 1 and 2?} 
We suspect that the main reason for such a behavior is that for large datasets
(Dataset~III-V, in our case) Samtools' SNP calling routine becomes more memory-
and time-demanding than the alignment with Bowtie.  Since its memory
requirements scale with input size, with sufficient input data Samtools swaps
data to disk, creating multiple (up to hundreds) temporary files while it sorts
the BAM file -- incidentally, these files can end up on shared network-accessed
file systems and thus generate large amounts of network traffic.
Table~\ref{table:5} shows the time required by Samtools to sort dataset~IV with
different amount of RAM available on a single HPC node. Though there is a large
deviation when using small amounts of RAM -- we believe due to other users' load
on the cluster network -- the trend quite clearly indicates that a relatively
large amount of RAM per node is needed in order to use Samtools effectively. In
contrast, in our tests Hadoop performed the SNP calling on much more economic
hardware, in linear time, with just 12GB of RAM per \textit{virtual} node.

\Change{%
Nevertheless, it is evident from the plot that the HPC method is always more
efficient in its use of CPU time than the Hadoop-based approach, and thus takes fewer computing
resources to perform the same amount of work.  However, Hadoop becomes
increasingly more efficient with respect to the HPC method as the dataset size
increases.
}
However, given that extrapolation to zero on the~$x$-axis estimates the ratio for the hypothetical
infinite dataset and yields the value~$1.70\pm0.01$, we can say that Hadoop, even if running in the virtualized environment of a private cloud running on modest hardware, is competitive with the HPC approach run on ``bare metal'' of modern hardware for datasets larger than $100$\,Gbases (dataset~IV), which is typical for human genome sequencing with a sequencing depth of about~$30x$.
\Change{%
In addition it is quite important to note that the price paid for
the efficiency in HPC's method is in overall runtime and in its inability to
finish the analysis on Dataset~V in a reasonable time (over 48 hours to Hadoop's 19.4
hours; see Table~\ref{table:4}).
}

\subsection*{Comparing the efficiency of network communication with the Hadoop and HPC approaches}

The network communication model for Hadoop has a major difference from the usual HPC cluster network architecture ($n$-tier tree) with NAS or NFS attached storages. The effective bandwidth of the Hadoop network increases with the cluster size~\cite{Sammer:2012}, opposite to that of the HPC network where cluster growth results in network saturation and performance depletion.
To study this phenomenon, we compared HPC and Hadoop network communication costs
\Fixme{(maybe ``network load''?)}
as a function of the number of nodes involved in a computation on a fixed dataset size (58~Gbases, dataset~IV).

Since the alignment process is embarrassingly parallel~-- the read-pairs are
independent of each other and thus can be aligned independently~-- it is
possible to distribute the work over more than one computational resource -- e.g., by
splitting the initial data into chunks to processing them independently. In this
approach, reducing the size of each data chunk reduces the size of the aligner
job,~$T_{alignment}$ but, at the same time, the more chunks almost
simultaneously have to be sent over the network, potentially causing traffic
jams, and therefore increasing the communication costs,~$T_{comm}$.

There are several program packages for short-read alignment with MPI support~\cite{pmap, gnumap} and almost linear scaling up to 32 nodes for pair-ended reads has been reported~\cite{Bozdag:2010cn}.
%
However, e.g., the Pmap package functions poorly for datasets larger than 20 Gbases.
\Fixme{(so, after reading the previous sentence, the logical question is ``why
not use gnumap?'')}
%
To circumvent this limitation we implemented a highly optimized Bash script
based on standard Unix utilities that makes very efficient use of the HPC
cluster network~\cite{code_repo_bash}; we then compared the network
performance with the standard Hadoop HDFS approach. We separated alignment
time, $T_{alignment}$, and communication time, $T_{comm}$, and plotted
the ratio $T_{alignment}/T_{comm}$ as a function of the reciprocal number of nodes~$1/N$
(Figure~\ref{fig:fig3}).  This measure is applicable to both HPC and Hadoop,
however, $T_{comm}$ has a different explanation. For Hadoop, the short reads in
FASTQ format have to be preprocessed (involving node communication~$T_{comm}$)
to be able to run in MR-fashion, while the data locality will be automatically
achieved during the data ingestion into the HDFS\@.  We rewrote the code for the
preprocessing stage for Crossbow to make it suitable for MR-style
parallelization.  For the HPC approach, $T_{comm}$ involves the chunks that are
sent from the sequence delivery location to the local node scratch disks where
the actual alignment happens, and the sending of the aligned SAM files back to
the delivery location over the network.
%\footnote{\url{http://samtools.sourceforge.net/SAMv1.pdf}}

@@

%The strategy named HPC SLURM\footnote{Simple Linux Utility for Resource Management}   is as follows: 
\textit{HPC results (open circles in Figure~\ref{fig:fig3})}: The data from the delivery location is split into chunks in parallel, which are simultaneously pushed to the local scratch disks of the nodes allocated by SLURM\@. One can see two distinct linear stretches. One stretch is for the range from 4 to about 12 nodes, and the another is from 12 up to 60. The former (horizontal stretch) is explained as for Hadoop~-- the more nodes involved, the faster the chunks are being distributed. The latter stretch with the positive slope could be explained as follows: In the region of about 12 nodes the network becomes saturated
%\footnote{The used storage at UPPMAX is a set of RAIDs (Redundant Array of Inexpensive Disks) with data striping, providing up to $80$Gbasesit/sec of outcoming traffic \COMMENT{'outgoing' or 'incoming' or just 'traffic'}.} 
and unable to pass more data in a time unit, while the alignment time is still proportional to the chunk size:~$T_{comm}\approx\mbox{const}, T_{alignment}\propto D\propto 1/N \rightarrow T_{alignment}/T_{comm}\propto 1/N$, i.e., a linear dependency, which can be observed in the plot. 
The transition area between the two modes is based on the fact that each hard disk drive on the local node can write the data at a speed of about 1Gbit/sec. Ten nodes will consume the data with the rate of~10Gbit/sec, which is the limiting speed for the standard 10Gbit Ethernet cable connecting the cluster's rack with the switch. The nodes are being allocated on the same rack, which is the default SLURM behavior.

@@

In this report we have described two approaches for high-performance analysis of
DNA sequencing data; one based on regular HPC using a batch system and one based
on Hadoop. We show that Hadoop installed on a private cloud is an appealing
solution both on terms of runtime performance and also in terms of usability. A
key result is the data size where Hadoop becomes an advantageous option as compared
to regular HPC batch systems.
In order to compare the two platforms we developed highly optimized pipelines
for both scenarios. We found that Hadoop excels over HPC when processing
datasets larger than 100\,Gbases. Also, Hadoop shows almost linear scalability for
this problem. Extrapolation to infinite dataset size reveals however that HPC
provides the results faster, given the same amount of resources as Hadoop on a
private cloud. \Fixme{But the HPC method tested ran on a single node so it
didn't incur distribution costs!}
